Visualizing Contingency Tables

### 2006-04-10 trunk/vcd/paper/hsc.Rnw
revision hsc
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@517 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

set.seed(myseed)
art_max <- coindep_test(art, n = 5000)
set.seed(myseed)
art_chisq <- coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))

@@

set.seed(myseed)
hec_max <- coindep_test(HairEyeColor, "Sex", n = 5000)

@@

\title*{Visualizing Multi-way Contingency Tables}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{David Meyer\inst{1}\and
Achim Zeileis\inst{2}\and
Kurt Hornik\inst{2}}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Department of Information Systems and Process Management
\and Department of Statistics and Mathematics\\Wirtschaftsuniversität
Wien\\Augasse 2--6, 1090 Vienna, Austria
\texttt{\{David.Meyer,Achim.Zeileis,Kurt.Hornik\}@wu-wien.ac.at}}
\maketitle

@@

All three graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \cite{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. 
\emph{Sieve plots} \cite{vcd:riedwyl+schuepbach:1994} 
are similar to mosaic plots, but the area of each
tile is proportional to the \emph{expected} cell entry, and each tile
is filled with a number of rectangles corresponding to the observed value.
An \emph{association plot} \cite{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

@@

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:hofmann:2001}, spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{itemize}

@@

\noindent The table relates the length of stay (in years) of 132
long-term schizophrenic patients in two London mental hospitals with
the freqency of visits. The length of stay (LOS) has been categorized into
2--9 years, 10--19 years, and more than 19 years. There are also three
categories for the visit frequency: regular (including
patients that were allowed to go home), less than monthly, and never.
Wing \cite{vcd:wing:1962} concludes from this data 
that the longer the length of stay in hospital, the less frequent the
visits, which can be seen from the column-standardized table 
(see Tab.~\ref{tab:colhospital}).

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|}
\hline
 & \multicolumn{3}{l|}{\textbf{Length of stay (in years)}}\\
\textbf{Visit frequency} &~~2--9 & ~~10--19 & ~~20+ \\
\hline
Regular & 0.74 & 0.36 & 0.10  \\
Less than monthly & 0.10 & 0.24 & 0.35  \\
Never & 0.16 & 0.40 & 0.55  \\\hline
$\Sigma$ & 1.00 & 1.00 & 1.00 \\
\hline
\end{tabular}
\caption{The \data{hospital} data, corrected for the column margin.}
\label{tab:colhospital}
\end{center}
\end{table}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<threedeebars,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
myhospital = t(hospital)[,3:1]
mydat = data.frame("Length of stay" = as.vector(row(myhospital)), 
                   "Visit frequency" = as.vector(col(myhospital)), 
                   "Number of patients" = as.vector(myhospital))
scatterplot3d(mydat, type = "h", pch = " ", lwd = 10, 
x.ticklabs = c("2-9","","10-19","","20+"), 
              y.ticklabs = c("Regular","","Less than monthly","","Never"), 
              xlab = "Length of stay (years)", ylab = "Visit frequency", zlab = "Number of patients",
              y.margin.add = 0.2,
              color = "black", box = F)
@
\caption{3D-bar chart for the \data{hospital} data.}
\label{fig:threedee}
\end{center}
\end{figure}

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a squared mosaic. The tiles are
obtained by recursive partitioning splits of a square.
In the following, we describe the main idea of mosaic plots, 
chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Fig.~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If causal
relationships between the variables are assumed, the determining variables
should be used first for splitting since they will dominate the
plot. Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They are best used to detect dependency
patterns, especially for ordinal variables. Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models is of primary concern.

@@

\subsection{A Note on Colors and Color Palettes}
\label{sec:colors}

@@

\begin{itemize}
\item Using high-chroma colors for large
  areas tend to produce after-image effects which can be
  distracting \cite{vcd:Ihaka:2003}.
\item Lighter colors tend to make areas look larger than darker
  colors, therefore using colors with unequal luminance values  makes it
  difficult to compare area sizes \cite{vcd:Cleveland+McGill:1983}.
\item Color palettes derived from
  non-uniform color spaces may contain unbalanced colors with respect to
  their colorfulness or brightness. When tiles are shaded using such a
  palette, some of them might appear more important than others in an
  uncontrolled way.
\end{itemize}

@@

Now, software implementations are often based on the more convenient
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$.}
\label{fig:colors}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<diverging,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
par(mfrow = c(1,1), mar = c(1,1,1,1), oma = c(0,0,0,0))
plot.new()
rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5))
rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = diverge_hsv(5))
rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
text(c(1:5/5 - 0.1), 0.11, c("(260, 100, 50)", "(260, 50, 70)", "(H, 0, 90)", "(0, 50, 70)", "(0, 100, 50)"))
text(c(1:5/5 - 0.1), 0.91, c("(240, 100, 100)", "(240, 50, 100)", "(H, 0, 100)", "(0, 50, 100)","(0, 100, 100)"))
@ 
\caption{Diverging color palettes for the HSV space (upper part) and
  the HCL space (lower part), ranging from blue over a neutral color to
  red. The triples indicate the settings for 
  the three dimensions: (Hue, Saturation, Value)
  in the upper part, and (Hue, Chroma, Luminance) in the lower part.}
\label{fig:diverging}
\end{center}
\end{figure}

@@

Using colors, even more complementary information can be visualized,
either by adding additional information, or by redundantly coding information
already visualized by the `raw' plot to support our perceptual system. %'
First, we consider the sieve plots. The density 
of the grid in the raw version implicitly gives us an 
idea of the residuals' %'
size, but since the plot does not include the density corresponding to zero
residuals (the null model) for comparison, we 
cannot easily assess whether there are more, of less counts in a cell 
than expected under the null hypothesis. It would help if we knew the 
sign of the residuals. 
Using color, we can add this information, for example using blue for
positive, red for negative, and gray for zero residuals. Figure~\ref{fig:colorsieve} 
demonstrates the effect of applying such a color shading to a sieve plot for the 
\data{hospital} data. Color clearly emphasizes the linear association
pattern in this data: Patients with a LOS from 2--9 years are
associated with regular visits, whereas patients with a LOS from
10--19 and more than 19 years are associated with a visit frequency
of less than monthly and no visits.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<sievecolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, shade = TRUE)
@ 
\caption{Sieve plot with color coding of the residuals.}
\label{fig:colorsieve}
\end{center}
\end{figure}

@@

In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5), 
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}

\noindent Clearly, the asymmetry for regular and never visited students,
and the pattern inversion for lengths of stay of 2--9 and more than 19
years are emphasized using the color shading.

@@

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
An important reason for using the unconditional limiting distribution
for the $X^2$ statistic from Equ.~\ref{eq:ChiSquare} was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from Equation~\ref{eq:H0},
using a permutation test is particularly intuitive due to the 
permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

Since the HCL space is three-dimensional and we only
used two `degrees of freedom' so far for coding information %'
(hue for the sign and a linear combination of chroma and luminance 
for the size of the residuals), we can add a third information to the plot.
For example, we can visualize the significance of some specified test statistic
(e.g., the $\chi^2$ test statistic) 
using darker (`uninteresting') colors for non-significant
results. These can again be derived using the interpolation procedure
described in the Sect.~\ref{sec:colors} by
restricting the luminance to a shorter range, ending with a ``darker'' value.

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Tab.~\ref{tab:arthritis} for the female patients):

@@

In Sect.~\ref{sec:twoway}, we have presented basic visualization methods
for two-way tables. In this section, we will show how these methods
extend to multi-way tables by applying them to `flat' representations,
and we will treat specialized displays for conditional independence models.

\subsection{Visualizing Flat-Tables}
\label{sec:flat}

>>>>

set.seed(myseed)
#art_max <- coindep_test(art, n = 5000)
uc_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Dept", n = 5000)

set.seed(myseed)
art_chisq <- coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))

@@

@@

\title*{Visualizing Contingency Tables}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{David Meyer\inst{1}\and
Achim Zeileis\inst{2}\and
Kurt Hornik\inst{2}}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Department of Information Systems and Operations
\and Department of Statistics and Mathematics\\Wirtschaftsuniversität
Wien\\Augasse 2--6, 1090 Vienna, Austria
\texttt{\{David.Meyer,Achim.Zeileis,Kurt.Hornik\}@wu-wien.ac.at}}
\maketitle

@@

All three graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \cite{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a rectangle. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. 
\emph{Sieve plots} \cite{vcd:riedwyl+schuepbach:1994} 
are similar to mosaic plots, but the area of each
tile is proportional to the \emph{expected} cell entry, and each tile
is filled with a number of rectangles corresponding to the observed value.
An \emph{association plot} \cite{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

@@

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:Hofmann:2001}, spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{itemize}

@@

\noindent The table relates the length of stay (in years) of 132
long-term schizophrenic patients in two London mental hospitals with
the frequency of visits (from relatives or friends). 
The length of stay (LOS) has been categorized into
2--9 years, 10--19 years, and more than 19 years. There are also three
categories for the visit frequency: regular (including
%FIXME: lookup regular in Wing-paper.
patients that were allowed to go home), less than monthly, and never.
Wing \cite{vcd:wing:1962} concludes from this data 
that the longer the length of stay in hospital, the less frequent the
visits, which can be seen from the column-standardized table 
(see Tab.~\ref{tab:colhospital}).

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|}
\hline
 & \multicolumn{3}{l|}{\textbf{Length of stay (in years)}}\\
\textbf{Visit frequency} &~~2--9 & ~~~~~~10--19 & ~~20+ \\
\hline
Regular & 0.74 & 0.36 & 0.10  \\
Less than monthly & 0.10 & 0.24 & 0.35  \\
Never & 0.16 & 0.40 & 0.55  \\\hline
$\Sigma$ & 1.00 & 1.00 & 1.00 \\
\hline
\end{tabular}
\caption{The \data{hospital} data, corrected for the column margin.}
\label{tab:colhospital}
\end{center}
\end{table}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<threedeebars,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
myhospital = t(hospital)[,3:1]
mydat = data.frame("Length of stay" = as.vector(row(myhospital)), 
                   "Visit frequency" = as.vector(col(myhospital)), 
                   "Number of patients" = as.vector(myhospital))
scatterplot3d(mydat, type = "h", pch = " ", lwd = 10, 
x.ticklabs = c("2-9","","10-19","","20+"), 
              y.ticklabs = c("Never","","Less than monthly","","Regular"), 
              xlab = "Length of stay (years)", ylab = "Visit frequency", zlab = "Number of patients",
              y.margin.add = 0.2,
              color = "black", box = F)
@
\caption{3D-bar chart for the \data{hospital} data.}
\label{fig:threedee}
\end{center}
\end{figure}

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Fig.~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If causal
relationships between the variables are assumed, the determining variables
should be used first for splitting since they will dominate the
plot. Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. However, for nominal variables, mosaic plots are
a better choice, especially when residual-based 
shadings are used (see next section), Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models is of primary concern.

@@

\subsection[A Note on Colors and Color Palettes]{A Note on Colors and Color Palettes\footnote{The electronic version of this article uses colors.}}
\label{sec:colors}

@@

\begin{itemize}
\item Using high-chroma colors for large
  areas tends to produce after-image effects which can be
  distracting \cite{vcd:Ihaka:2003}.
\item Lighter colors tend to make areas look larger than darker
  colors, therefore using colors with unequal luminance values  makes it
  difficult to compare area sizes \cite{vcd:Cleveland+McGill:1983}.
\item Color palettes derived from
  non-uniform color spaces may contain unbalanced colors with respect to
  their colorfulness or brightness. When tiles are shaded using such a
  palette, some of them might appear more important than others in an
  uncontrolled way.
\end{itemize}

@@

Now, software implementations are often based on the more convenient
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e. appear identical.}
\label{fig:colors}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<diverging,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
par(mfrow = c(1,1), mar = c(1,1,1,1), oma = c(0,0,0,0))
plot.new()
rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5))
rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = diverge_hsv(5))
rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
text(c(1:5/5 - 0.1), 0.11, c("(260, 100, 50)", "(260, 50, 70)", "(H, 0, 90)", "(0, 50, 70)", "(0, 100, 50)"))
text(c(1:5/5 - 0.1), 0.91, c("(240, 100, 100)", "(240, 50, 100)", "(H, 0, 100)", "(0, 50, 100)","(0, 100, 100)"))
@ 
\caption{Diverging color palettes for the HSV space (upper part) and
  the HCL space (lower part), ranging from blue over a neutral color to
  red. The triples indicate the settings for 
  the three dimensions: (Hue, Saturation, Value)
  in the upper part, and (Hue, Chroma, Luminance) in the lower part.
  In a monochrome version of the paper, the right and left hand
  sides of the HCL colors palette appear identical. }
\label{fig:diverging}
\end{center}
\end{figure}

@@

Using colors, even more complementary information can be visualized,
either by adding additional information, or by redundantly coding information
already visualized by the `raw' plot to support our perceptual system. %'
First, we consider the sieve plots. The density 
of the grid in the raw version implicitly gives us an 
idea of the residuals' %'
size, but since the plot does not include the density corresponding to zero
residuals (the null model) for comparison, we 
cannot easily assess whether there are more, of less counts in a cell 
than expected under the null hypothesis. 
%It would help if we knew the sign of the residuals. 
Using color, we can add the sign information, for example using blue for
positive, red for negative, and gray for zero residuals.

% Figure~\ref{fig:colorsieve} 
% demonstrates the effect of applying such a color shading to a sieve plot for the 
% \data{hospital} data. Color clearly emphasizes the linear association
% pattern in this data: Patients with a LOS from 2--9 years are
% associated with regular visits, whereas patients with a LOS from
% 10--19 and more than 19 years are associated with a visit frequency
% of less than monthly and no visits.

% \setkeys{Gin}{width=0.7\textwidth}
% \begin{figure}
% \begin{center}
% <<sievecolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
% sieve(t(hospital), split = TRUE, shade = TRUE)
% @ 
% \caption{Sieve plot with color coding of the residuals.}
% \label{fig:colorsieve}
% \end{center}
% \end{figure}

@@

In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5), 
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}
Clearly, the asymmetry for regular and never visited students,
and the pattern inversion for lengths of stay of 2--9 and more than 19
years are emphasized using the color shading.

@@

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
An important reason for using the unconditional limiting distribution
for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from (\ref{eq:H0}),
using a permutation test is particularly intuitive due to the 
permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

Since the HCL space is three-dimensional and we only
used two `degrees of freedom' so far for coding information %'
(hue for the sign and a linear combination of chroma and luminance 
for the size of the residuals), we can add a third piece of information to the plot.
For example, we can visualize the significance of some specified test statistic
(e.g., the $\chi^2$ test statistic) 
using darker (`uninteresting') colors for non-significant
results. These can again be derived using the interpolation procedure
described in Sect.~\ref{sec:colors} by
restricting the luminance to a shorter range, ending with a ``darker'' value.

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Tab.~\ref{tab:arthritis} for the female patients):

@@

In Sect.~\ref{sec:twoway}, we have presented basic displays
for two-way tables, based on the visualization of information on 
the tables' cells, arranged in rectangular form. For multi-way tables,
mosaic plots can directly be used by simply adding further splits for each 
additional variable. For sieve and association plots,
we apply the basic idea of mosaic plots to the table itself,
i.e., simply nest the variables into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation of the multi-way table that can be visualized in %'
ways similar to a two-dimensional table.

In the following, we will first treat specialized displays for
explorative purposes, followed by model-based methods for
conditional independence models.

\subsection{Explorative Visualization Techniques}
\label{sec:flat}

As an example, consider the well-known UCB admissions data 
\cite{vcd:Bickel+Hammel+O'Connell:1975}
on applicants to graduate school at the University of California in Berkeley for
the six largest departments in 1973 classified by admission and
gender. The `flattened' contingency table, putting department into the
columns and admit---nested in gender---into the rows, is given by
Tab.~\ref{tab:ucb}. The table aggregated over all
departments gives wrong evidence of gender bias, 
i.e., higher admission rates for male students.

A first step in the explorative analysis of more complex tables is to
get a quick overview on the data. For that, 
all basic plots can be combined in pairwise displays,
arranged in in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Fig.~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and also varies given department.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<pairs,fig=TRUE,echo=FALSE,height=8,width=8,results=hide>>=
pairs(UCBAdmissions)
@
\caption{Pairs plot for the \data{HairEyeColor} data.}
\label{fig:pairs}
\end{center}
\end{figure}
